---
title: "Big Data and Consumer Analytics Assignment 1"
author: "Student ID: 201075916"
date: "5 March 2019"
output: pdf_document
---


## Data Exploration 

In order to perform any analysis, it is crucial to identify how the files can be linked, in this case the variables "upc" and "store".  This can be seen in the code and figures below. The common fields are: the "upc" and "store" variables. 

```{r}
# test for package existance and install
if (!is.element("tidyverse", installed.packages()))
  install.packages("tidyverse", dep = T)
if (!is.element("nycflights13", installed.packages()))
  install.packages("nycflights13", dep = T)
# load into the R session
library(tidyverse)
library(nycflights13)


#load the data as tibble
library(tidyverse)
cl <- as_tibble(read.csv("dunnhumby/dh_causal_lookup.csv"))
pl <- as_tibble(read.csv("dunnhumby/dh_product_lookup.csv"))
sl <- as_tibble(read.csv("dunnhumby/dh_store_lookup.csv"))
tr <- as_tibble(read.csv("dunnhumby/dh_transactions.csv"))

#inspect the data to find the common field
head(cl)
head(pl)
head(sl)
head(tr)

```

## Q1: What is the household penetration of `PRIVATE LABEL THIN SPAGHETTI`? That is, out of all customers purchasing Pasta, what percent purchase this brand?

In order to answer this task, the number of people that have bought this specific brand, i.e. "Private Label This Spaghetti" must be calculated along with the number of people that have bought the respective commodity, in this case "Pasta". In order to do this, the file of transactions was joined with the product file through the variable "upc". The variables necessary were then kept through the "select" command. Then, the two datasets that were created were filtered to keep records only with "Private Label This Spaghetti" and "Pasta" respectively.Finally, the two datasets were saved in two dataframes, from which the number of unique values was calculated, in order to estimate the percentage of people that bought that specific pasta. The code that was used in this part is visible in figure 2 and in figure 3 a graph showing the household penetration can be seen. The final percentage is 22.07% of household penetration of "Private Label This Spaghetti". 

```{r}

#begin with the transaction file
tr %>%
  #join the files of transaction and product look up
  left_join(pl) %>% 
  #keep the variables that are of our interest
  select("upc", "geography", "household","basket", "product_description")%>% 
  #select only the purchases with the desired label
  filter(product_description == "PRIVATE LABEL THIN SPAGHETTI")%>% 
  #get the count by household and save it in a table
  count(household) -> n.hhpen 

tr %>%
  #join the files of transaction and product look up
  left_join(pl) %>% 
  #keep the variables that are of our interest
  select("upc", "geography", "household","basket", "product_description", "commodity")%>%
  #select only the purchases with the desired commodity
  filter(commodity == "pasta")%>% 
  #get the count by household and save it in a table
  count(household) -> n.hhpen2 

#in order to calculate the household penetration, we need to calculate the percentage of 
#people who bought the label "Private label thin spaghetti" and 
#the people who bought "pasta"(commodity) in general

#divide by length of each table, which will be the unique households in each case
hh_pen <- length(n.hhpen$household)*100/length(unique(n.hhpen2$household))

#Therefore, the answer to the first question is 22.065% or 22.07%

```


```{r, echo= FALSE}

Household_penetration<-"Household penetration"
Percentage<- round(hh_pen,2)
graph<- data.frame(Household_penetration, Percentage)
head(graph)

ggplot(graph, mapping=aes(Household_penetration, Percentage, label="22.07%"))+
  geom_bar(stat="identity", width = 0.30, fill= "steelblue")+
  theme(axis.text.x=element_blank())+
  coord_cartesian(ylim = c(0, 25))+
  geom_text(size = 3, position = position_stack(vjust = 0.5))+
  ggtitle("Household Penetration of Private Label Thin Spaghetti")
```
Figure 1. Graph for household penetration of "Private Label Thin Spaghetti" (RStudio, 2019).  

## Q2: How does the household penetration of the product PRIVATE LABEL THIN SPAGHETTI vary within the two regions, relative to the sales of Pasta in each region?

For this task, the previous code with the variable of "geography" was replicated, so that the two regions could be studied separately. Then the datasets that were created were filtered for each region in order to calculate the percentage. This was conducted twice, once for the "Private Label This Spaghetti" and once for the commodity "Pasta". The household penetration for each region is 18.32% and 27.01% respectively.  
```{r}
#begin with the transaction file
tr %>%
  #join the files of transaction and product look up
  left_join(pl) %>% 
  #keep the variables that are of our interest
  select("geography", "household", "product_description")%>% 
  #filter for product
  filter(product_description == "PRIVATE LABEL THIN SPAGHETTI")%>%
  #get the unique values of households by geography
  distinct(household, geography)%>%
  #get the count of households for each geography
  count(geography)->product_region

#the same for the commodity pasta
tr %>%
  #join the files of transaction and product look up
  left_join(pl) %>%
  #keep the variables that are of our interest
  dplyr::select("upc", "geography", "household", "product_description", "commodity")%>%
  #filter for commodity 
  filter(commodity == "pasta")%>% 
  #get the unique values of households by geography
  distinct(household, geography)%>%
  #get the count of households for each geography
  count(geography)->commodity_region

#household penetration for region 1, in row 1 and column 2 for each dataset
hh_pen_1<-as.numeric((product_region[1,2])*100/(commodity_region[1,2]))
print(hh_pen_1)

#household penetration for region 2, in row 2 and column 2 for each dataset
hh_pen_2<-as.numeric((product_region[2,2])*100/(commodity_region[2,2]))
print(hh_pen_2)

```


```{r, echo = FALSE}
Regions<- c("Region 1", "Region 2")
Percentage<-c(round(hh_pen_1,2), round(hh_pen_2,2) ) 
graph<- data.frame(Regions, Percentage)
head(graph)

ggplot(graph, mapping=aes(Regions, Percentage, label=c("18.32%", "27.01%")))+
  geom_bar(stat="identity", width = 0.30, fill= c("steelblue","deeppink3"))+
  coord_cartesian(ylim = c(0, 30))+
  geom_text(size = 3, position = position_stack(vjust = 0.5))+
  ggtitle("Variance of household penetration of product in the two regions")
```
Figure 2. Graph for household penetration of "PRIVATE LABEL THIN SPAGHETTI" in each region (RStudio, 2019) 

## Q3: For which category of products (pasta, pasta sauces, syrups and pancake mixes) does theprovision of coupons appear to have a positive impact?

This task will be solved with the creation of a function for each commodity in order to minimise the effort. 

It is of the utmost importance to consider how the provision will be calculated, i.e., which households. The provision of coupons will be calculated based on the households the first purchase of the commodity was with the use of a coupon, and purchased the commodity more than once. First of all, a dataframe will be created that has the households that only bought the desired commodity with a coupon. A second dataframe was created that was filtered for the commodity, but had the transactions arranged by day(earliest to latest) and use of coupon first. Then, with a "for" loop, the households from the first datasets were used to filter the second dataset and check whether the first purchase was done with the use of coupon and whether there were more purchases. The impact was calculated by the percentage of people that bought the commodity the first time with a coupon and then repurchased divided by the number of people that bought the commodity with a coupon in general. 
```{r}

## Creation of function

#find the households that bought with a coupon
impact<-function(commod){
  #join product info data and transaction data
  tr%>%
    left_join(pl)%>%
    #filter for use of coupon
    filter(coupon == 1)%>% 
    #select the variables of our interest
    dplyr::select("upc",  "household","day", "product_description", "commodity","coupon")%>%
    #filter for commodity
    filter(commodity == commod)-> r1 
  
  #find number of households that bought a product with a coupon
  n.hh<- as.data.frame(unique(r1$household))
  
  #Examine the transactions of each household individually by day and use of coupon
  
  #start with the transaction dataset
  tr%>%
    #join the data with product_lookup file
    left_join(pl)%>% 
    #select variables of our interest
    dplyr::select("upc",  "household","day", "product_description", "commodity","coupon")%>%
    #filter for the specific commodity
    filter(commodity == commod)%>%
    #arrange the data from earliest day to latest and from use of coupon to no use of coupon
    arrange(day, -coupon)->r2 
  
 
  #create an empty vector where the households will be appended later
  r<-vector("numeric")
  
  #create a "for" loop in order to go through all the houses that have 
  #made a purchase with a coupon
  for (row in 1:nrow(n.hh)){
    #start with the 
    r2%>%
      #filter for the household from the dataset of households that purchased the commodity with a coupon
      filter(household == n.hh[row, 1])%>%
      #arrange the data for the particular household by day and use of coupon (same as before)
      arrange(day,-coupon)%>%
      #select the variable coupon
      select(coupon)%>%
      #unlist the data
      unlist->tmp
    #create "if" statement to check whether 
    #1. the FIRST purchase was made with a coupon
    #2. there were more than one purchases
    if (tmp[1]== 1 & length(tmp)>1)
      #if the above statement is true, append the household to the vector r
      r<- append(r,n.hh[row,1])
  }
  #Number of households that fulfill the "if" statement
  imp_coup<- length(r)
  #percentage in order to determine the impact of coupons
  #percentage of number of households that bought the commodity with a coupon first and then repurchased 
  #divided by the number of people that bought the commodity with a coupon in general 
  impact_comm<-imp_coup*100/length(unique(r1$household))
  print(impact_comm)
}

impact_pasta<- impact("pasta")
impact_pasta_sauce<-impact("pasta sauce")
impact_pancake_mixes<-impact("pancake mixes")
impact_syrups<-impact("syrups")
```

The coupons have the highest impact on the commodity of "Pasta sauces" with 29.53% of people purchasing the commodity again, then "Pancake mixes" (27.34%), then "Syrups" (26.74%) and finally "Pasta" with 25.34%. The graph below illustrates these values for each commodity. 
```{r, echo=FALSE}
Commodity<- c("Pasta", "Pasta sauce", "Pancake mixes", "Syrups")
Percentage<-c(round(impact_pasta,2), round(impact_pasta_sauce,2), round(impact_pancake_mixes,2), round(impact_syrups,2)) 
graph<- data.frame(Commodity, Percentage)
head(graph)

ggplot(graph, mapping=aes(Commodity, Percentage, label=c("25.34%", "29.53%", "27.34%", "26.74%")))+
  geom_bar(stat="identity", width = 0.50, fill= c("cadetblue","cadetblue1", "cadetblue2", "cadetblue4"))+
  coord_cartesian(ylim = c(0, 30))+
  geom_text(size = 3, position = position_stack(vjust = 0.5))+
  ggtitle("Impact of coupons on each commodity") 
```
Figure 3. Graph  of impact of coupons on sales of each commodity (RStudio, 2019) 

## References 

Heilman, C. 2019. Dunnhumby Carbo-Loading: A Relational Database. Dunnhumby. [Online]. [Accessed 21 February 2019]. Available from: https://www.dunnhumby.com/careers/engineering/sourcefiles 

Comber, L. 2019.Lectures and Practicals from Big Data and Consumer Analytics.[HTMLs accessed through Minerva]. GEOG5917M Big Data and Consumer Analytics.University of Leeds